Perceptions of Inequality and Meritocracy: Their Interplay in Shaping Preferences for Market Justice in Chile (2016-2023)

### 2025-07-11 paper.qmd
Update carpeta resubmission y varios (archivos no renderizados)

---
title: "Perceptions of Inequality and Meritocracy: Their Interplay in Shaping Preferences for Market Justice in Chile (2016-2023)"
format:
  html: 
    code-fold: true
    html-math-method: katex
    number-sections: true
  docx:
    number-sections: true
  pdf: 
    geometry: margin=2cm
    template-partials: 
      - title.tex
    keep-tex: true
    include-in-header:
      text: |
        \usepackage[noblocks]{authblk}
        \renewcommand*{\Authsep}{, }
        \renewcommand*{\Authand}{, }
        \renewcommand*{\Authands}{, }
        \renewcommand\Affilfont{\small}
    number-sections: true
editor: source
bibliography: "input/bib/merit-edjust.bib"
# csl: "input/bib/apa6.csl"
csl: "input/bib/harvard.csl"
author:
  - name: Juan Carlos Castillo
    corresponding: true
    orcid: 0000-0003-1265-7854
    email: juancastillov@uchile.cl
    affiliations:
      - ref: 1
      - ref: 2
      - ref: 3
  - name: Andreas Laffert
    orcid: 0000-0002-9008-2454
    email: alaffertt@estudiante.uc.cl 
    affiliations:
      - ref: 4
  - name: Kevin Carrasco
    orcid: 0000-0002-9348-8361
    email: kevin.carrasco@ug.uchile.cl
    affiliations:
      - ref: 2
  - name: Julio Iturra
    orcid: 0000-0003-1993-3479
    email: julioiturrasanhueza@gmail.com 
    affiliations:
      - ref: 5     
affiliations: 
  - id: 1
    name: Departamento de Sociología, Universidad de Chile
  - id: 2
    name: Centro de estudios del conflicto y cohesión social (COES)
  - id: 3
    name: Núcleo milenio de desigualdades y oportunidades digitales (NUDOS)
  - id: 4
    name: Instituto de Sociología, Pontificia Universidad Católica de Chile
  - id: 5
    name: International Graduate School of Social Sciencies (BIGSSS), University of Bremen, Germany
citeproc: true
abstract: | 
  This study investigates the relationship between perceptions of economic inequality, meritocratic beliefs, and preferences for market justice in Chile between 2016 and 2023. Using six waves of panel data from the Chilean Longitudinal Social Survey - ELSOC ($N_{observations}$ = 8,643; $N_{individuals}$ = 1,687), the analysis examines how subjective assessments of inequality shape attitudes toward the role of merit in access to key social services such as healthcare, education, and pensions. Results indicate that rising perceptions of inequality are associated with greater support for redistributive policies; however, individuals with strong meritocratic convictions are more likely to legitimize existing disparities. The study also considers the influence of major social movements during this period, which appear to have reshaped public discourse and perceptions of fairness. These findings contribute to a deeper understanding of how beliefs about justice and equity evolve in contexts marked by persistent inequality and entrenched market-oriented frameworks.
  \newline
  **Keywords**: Economic inequality, meritocracy, market justice, Chile, public preferences, inequality perception
link-citations: true
linestretch: 1.15       
mainfont: Times New Roman
fontsize: 12pt          
colorlinks: true
fig-height: 4
fig-width: 7.5
---

>>>>

---
title: "Perceptions of Inequality and Meritocracy: Their Interplay in Shaping Preferences for Market Justice in Chile (2016-2023)"
format:
  html: 
    code-fold: true
    html-math-method: katex
    number-sections: true
  docx:
    number-sections: true
  pdf: 
    geometry: margin=2cm
    template-partials: 
      - title.tex
    keep-tex: true
    include-in-header:
      text: |
        \usepackage[noblocks]{authblk}
        \renewcommand*{\Authsep}{, }
        \renewcommand*{\Authand}{, }
        \renewcommand*{\Authands}{, }
        \renewcommand\Affilfont{\small}
    number-sections: true
editor: source
bibliography: "input/bib/merit-edjust.bib"
# csl: "input/bib/apa6.csl"
csl: "input/bib/harvard.csl"
author:
  - name: Juan Carlos Castillo
    corresponding: true
    orcid: 0000-0003-1265-7854
    email: juancastillov@uchile.cl
    affiliations:
      - ref: 1
      - ref: 2
      - ref: 3
  - name: Andreas Laffert
    orcid: 0000-0002-9008-2454
    email: alaffertt@estudiante.uc.cl 
    affiliations:
      - ref: 4
  - name: Kevin Carrasco
    orcid: 0000-0002-9348-8361
    email: kevin.carrasco@ug.uchile.cl
    affiliations:
      - ref: 2
  - name: Julio Iturra-Sanhueza
    orcid: 0000-0003-1993-3479
    email: julioiturrasanhueza@gmail.com 
    affiliations:
      - ref: 5     
affiliations: 
  - id: 1
    name: Departamento de Sociología, Universidad de Chile
  - id: 2
    name: Centro de estudios del conflicto y cohesión social (COES)
  - id: 3
    name: Núcleo milenio de desigualdades y oportunidades digitales (NUDOS)
  - id: 4
    name: Instituto de Sociología, Pontificia Universidad Católica de Chile
  - id: 5
    name: International Graduate School of Social Sciencies (BIGSSS), University of Bremen, Germany
citeproc: true
abstract: | 
  This study investigates the relationship between perceptions of economic inequality, meritocratic perceptions, and preferences for market justice in Chile between 2016 and 2023. Using six waves of panel data from the Chilean Longitudinal Social Survey - ELSOC ($N_{observations}$ = 8,643; $N_{individuals}$ = 1,687), the analysis examines how subjective assessments of inequality shape attitudes toward the role of merit in access to key social services such as healthcare, education, and pensions. Results indicate that rising perceptions of inequality are associated with lower support for market justice; however, individuals with strong meritocratic convictions are more likely to legitimize existing disparities. The study also considers the influence of major social movements during this period, which appear to have reshaped public discourse and perceptions of fairness. These findings contribute to a deeper understanding of how beliefs about justice and equity evolve in contexts marked by persistent inequality and entrenched market-oriented frameworks.
  \newline
  **Keywords**: Economic inequality, meritocracy, market justice, Chile, public preferences, inequality perception
link-citations: true
linestretch: 1.15       
mainfont: Times New Roman
fontsize: 12pt          
colorlinks: true
fig-height: 4
fig-width: 7.5
---

### 2025-07-11 resubmission/02-antecedents.qmd
Update carpeta resubmission y varios (archivos no renderizados)

The study of market justice preferences has increasingly focused on how they are shaped by individuals' socioeconomic position, normative beliefs, and the institutional context in which they are embedded. Across the literature, there is empirical evidence suggesting that individuals in more advantaged socioeconomic positions---those with higher occupational class, income, and education---are more likely to support market justice principles [@svallfors_political_2007; @koos_moral_2019]. This tendency reflects not only material self-interest but also a broader moral economy, in which winners of the market system internalize justifications for the status quo. At the same time, political ideology also plays a role ---such as economically conservative values ---  where more right-wing individuals show higher support for meritocracy and more skepticism towards redistribution. This is particularly salient in countries with more restricted public provision of social services. For example, Castillo et al. [-@castillo_socialization_2024] in Chile and, Lee and Stacey [-@lee_fairness_2023] in Australia show that those with right-leaning individuals are more supportive of market-based distribution of welfare. Beyond individual characteristics, country-level institutions also play a central role. In liberal welfare regimes like those of the United States or the United Kingdom, market justice preferences are more widespread, while in coordinated or social-democratic regimes---such as Sweden or Germany---citizens are generally more critical of market-based inequalities [@immergut_it_2020; @lindh_public_2015]. These findings suggest that market justice is also shaped contextually, which is represented by the development of welfare institutions, in line with the policy-feedbacks literature [@campbell_policy_2012; @pierson_when_1993].

@@

Perceptions of inequality have been associated with attitudes about market justice [@lindh_public_2015; @kluegel_social_1995a]. Research indicates that lower perceived inequality can reinforce support for market-based distributive arrangements by suggesting that the system is fair and that outcomes reflect effort and ability [@kuhn_eye_2011]. In contrast, when inequality is perceived as excessive or structurally determined, individuals could question the legitimacy of market justice and become more supportive of redistributive policies [@castillo_perception_2022; @garcia-sanchez_vicious_2019].

@@

Second, higher perceived meritocracy is expected to be positively associated with market justice preferences. Individuals who believe that effort and talent primarily determine success are more likely to justify unequal outcomes and endorse the notion that markets allocate resources fairly according to individual merit. This aligns with the idea that meritocratic beliefs serve as a moral framework that legitimizes market-based inequalities, as individuals perceive the system as just when they believe that rewards are based on individual merit. This is particularly relevant in contexts where neoliberal ideologies dominate, as they often emphasize individual responsibility and competition as the basis for social order.

Third, we propose that perceptions of meritocracy and perceptions of economic inequality interact in shaping market justice attitudes. Specifically, we argue that the legitimizing effect of perceived meritocracy on market justice preferences is moderated by perceived economic inequality: when perceived inequality is low, the positive association between meritocratic beliefs and market justice preferences will be stronger. However, when perceived inequality is high, this association will weaken, as greater awareness of large economic gaps may challenge the view that outcomes are purely merit-based.

@@

$H2$: Higher meritocratic beliefs are associated with higher market justice preferences.

$H3$: The positive association between meritocratic beliefs and market justice preferences is moderated by perceived economic inequality; specifically, this association is weaker when perceived economic inequality is high.

>>>>

The study of market justice preferences has increasingly focused on how they are shaped by individuals' socioeconomic position, normative beliefs, and the institutional context in which they are embedded. Across the literature, there is empirical evidence suggesting that individuals in more advantaged socioeconomic positions---those with higher occupational class, income, and education---are more likely to support market justice principles [@svallfors_political_2007; @koos_moral_2019]. This tendency reflects not only material self-interest but also a broader moral economy, in which winners of the market system internalize justifications for the status quo. At the same time, political ideology also plays a role ---such as economically conservative values ---  where more right-wing individuals show higher support for meritocracy and more skepticism towards redistribution. This is particularly salient in countries with more restricted public provision of social services. For example, Castillo et al. [-@castillo_socialization_2024] in Chile and, Lee and Stacey [-@lee_fairness_2023] in Australia show that right-wing individuals are more supportive of market-based distribution of welfare. Beyond individual characteristics, country-level institutions also play a central role. In liberal welfare regimes like those of the United States or the United Kingdom, market justice preferences are more widespread, while in coordinated or social-democratic regimes---such as Sweden or Germany---citizens are generally more critical of market-based inequalities [@immergut_it_2020; @lindh_public_2015]. These findings suggest that market justice is also shaped contextually, which is represented by the development of welfare institutions, in line with the policy-feedbacks literature [@campbell_policy_2012; @pierson_when_1993].

@@

Perceptions of inequality have been associated with attitudes about market justice [@lindh_public_2015; @kluegel_social_1995a]. Research indicates that lower perceived inequality can reinforce support for market-based distributive arrangements by suggesting that the system is fair and that outcomes reflect effort and ability [@kuhn_eye_2011]. In contrast, when inequality is perceived as excessive or structurally determined, individuals could question the legitimacy of market justice and become more supportive of redistributive policies [@castillo_perception_2022; @garcia-sanchez_vicious_2019].ƒ

@@

Second, higher perceived meritocracy is expected to be positively associated with market justice preferences. Individuals who believe that effort and talent primarily determine success are more likely to justify unequal outcomes and endorse the notion that markets allocate resources fairly according to individual merit. This aligns with the idea that meritocratic perceptions serve as a moral framework that legitimizes market-based inequalities, as individuals perceive the system as just when they believe that rewards are based on individual merit. This is particularly relevant in contexts where neoliberal ideologies dominate, as they often emphasize individual responsibility and competition as the basis for social order.

Third, we propose that perceptions of meritocracy and perceptions of economic inequality interact in shaping market justice attitudes. Specifically, we argue that the legitimizing effect of perceived meritocracy on market justice preferences is moderated by perceived economic inequality: when perceived inequality is low, the positive association between meritocratic perceptions and market justice preferences will be stronger. However, when perceived inequality is high, this association will weaken, as greater awareness of large economic gaps may challenge the view that outcomes are purely merit-based.

@@

$H2$: Higher meritocratic perceptions are associated with higher market justice preferences.

$H3$: The positive association between meritocratic perceptions and market justice preferences is moderated by perceived economic inequality; specifically, this association is weaker when perceived economic inequality is high.

### 2025-07-11 resubmission/03-methods.qmd
Update carpeta resubmission y varios (archivos no renderizados)

---
author: "Equipo EDUMER"
bibliography: "input/bib/merit-edjust.bib"
csl: "input/bib/apa6.csl"
---

@@

pacman::p_load(tidyverse, 
               sjmisc, 
               sjPlot, 
               lme4, 
               here, 
               performance,
               marginaleffects,
               texreg, 
               ggdist,
               kableExtra,
               ggalluvial, 
               shadowtext,
               MetBrewer,
               patchwork,
               sjlabelled,
               summarytools)

>>>>

---
author: "Equipo EDUMER"
bibliography: "../input/bib/merit-edjust.bib"
csl: "../input/bib/apa6.csl"
---

@@

pacman::p_load(tidyverse, 
               sjmisc, 
               sjPlot, 
               lme4, 
               here, 
               performance,
               influence.ME,
               marginaleffects,
               #MLMusingR,
               texreg, 
               ggdist,
               misty,
               kableExtra,
               ggalluvial, 
               shadowtext,
               MetBrewer,
               patchwork,
               sjlabelled,
               summarytools)
